---
title: "curatedCRCData: Clinically Annotated Data for the Colorectal Cancer Transcriptome"
author: "Princy Parsana, Markus Riester, Curtis Huttenhower, Levi Waldron"
date: "2013"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{curatedCRCData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r preliminaries, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(sva)
library(xtable)
library(Biobase)
```

# curatedCRCData: Clinically Annotated Data for the Colorectal Cancer Transcriptome

This package represents a manually curated data collection for gene expression
meta-analysis of patients with colorectal cancer. This resource provides
uniformly prepared microarray data with curated and documented clinical
metadata. It allows a computational user to efficiently identify studies and
patient subgroups of interest for analysis and to run such analyses
immediately without the challenges posed by harmonizing heterogeneous
microarray technologies, study designs, expression data processing methods,
and clinical data formats.

In this vignette, we give a short tour of the package and will show how to use
it efficiently.

# Data access and caching

As of version 2.46, the datasets are no longer stored inside the package:
they are hosted on [Zenodo](https://zenodo.org/) and downloaded
individually the first time they are requested, then cached locally with
`r BiocStyle::Biocpkg("BiocFileCache")` so later uses need no network
access. The cache lives in a package-specific directory,
`tools::R_user_dir("curatedCRCData", "cache")`; to reclaim disk space,
delete that directory or manage it with *BiocFileCache*.

The legacy `data(TCGA.COAD_eset)` syntax still works (it downloads
through the same cache) but is deprecated and will be removed in a
future release; please use the `curatedCRCData()` function shown below.

Throughout this vignette we set `test = TRUE` (and `test.mode <- TRUE`),
which loads small offline subsets of each dataset that are bundled with
the package, so that the vignette builds without downloading data. For
real analyses, omit these settings to work with the full datasets.

# Load data sets
Loading a single dataset is very easy. First we load the package:
```{r example1, message=FALSE, warning=FALSE}
library(curatedCRCData)
```

To get a listing of all the datasets, call `curatedCRCData()` with no
arguments:
```{r example1tcgastep2}
curatedCRCData()
```

Now to load a single dataset (drop `test = TRUE` to download the full
dataset):
```{r example1_load_tcga}
TCGA.COAD_eset <- curatedCRCData("TCGA.COAD_eset", test = TRUE)
TCGA.COAD_eset
```

The datasets are provided as Bioconductor `ExpressionSet` objects and
we refer to the Bioconductor documentation for users unfamiliar with this data
structure.

# Load datasets based on rules
For a meta-analysis, we typically want to filter datasets and patients to
get a population of patients we are interested in. We provide a short but
powerful R script that does the filtering and provides the data as a list of
`ExpressionSet` objects. One can use this script within R by first sourcing a config
file which specifies the filters, like the minimum numbers of patients in each
dataset. It is also possible to filter samples by annotation, for example to remove
early stage and normal samples. 

```{r example2loadstep1}
source(system.file("extdata",
"patientselection_all.config",package="curatedCRCData"))
test.mode <- TRUE  # use the bundled offline subsets in this vignette
ls()
```

See what the values of these variables we have loaded are. The
variable names are fairly descriptive, but note that `rule.1` is a
character vector of length 2, where the first entry is the name of a
clinical data variable, and the second entry is a Regular Expression
providing a requirement for that variable. Any number of rules can be
added, with increasing identifiers, e.g. `rule.2`, `rule.3`, etc.

Here strict.checking is FALSE, meaning that samples not annotated for
the variables in these rules are allowed to pass the filter. If
strict.checking == TRUE, samples missing this annotation will be
removed.

```{r showls}
sapply(ls(), get)
```

Now that we have defined the sample filter, we create a list of
`ExpressionSet` objects by sourcing the `createEsetList.R` file:

```{r example2loadstep2}
source(system.file("extdata", "createEsetList.R", package = "curatedCRCData"))
```

It is also possible to run the script from the command line and then load the
R data file within R:
```
R --vanilla "--args patientselection.config crc.eset.rda tmp.log"  < createEsetList.R 
```

Now we have `r length(esets)` datasets with samples that passed our filter in a list of
`ExpressionSet` objects called `esets`:

```{r example2loadstep3}
names(esets)
```

# Non-unique gene symbols

In the standard version of curatedCRCData (the version available on
Bioconductor), we collapse manufacturer probesets to official HGNC symbols
using the Biomart database. Some probesets are mapped to multiple HGNC symbols
in this database. For these probesets, we provide all the symbols. For example 
`220159_at` maps to *ABCA11P* and *ZNF721* and we
provide `ABCA11P///ZNF721` as probeset name. If you have an array of
gene symbols for which you want to access the expression data, "ABCA11P" would
not be found in curatedCRCData in this example. The following function
will create a new ExpressionSet in which both *ZNF721* and
*ABCA11P* are features with identical expression data:

```{r expand}
expandProbesets <- function (eset, sep = "///") 
{
    x <- lapply(featureNames(eset), function(x) strsplit(x, sep)[[1]])
    eset <- eset[order(sapply(x, length)), ]
    x <- lapply(featureNames(eset), function(x) strsplit(x, sep)[[1]])
    idx <- unlist(sapply(1:length(x), function(i) rep(i, length(x[[i]]))))
    xx <- !duplicated(unlist(x))
    idx <- idx[xx]
    x <- unlist(x)[xx]
    eset <- eset[idx, ]
    featureNames(eset) <- x
    eset
}

X <- TCGA.COAD_eset[head(grep("AA", featureNames(TCGA.COAD_eset))),]
exprs(X)[,1:3]
exprs(expandProbesets(X))[,1:3]
```

# Appendix

## Available Clinical Characteristics

```{r heatmap, echo=FALSE, fig.cap="Available clinical annotation. This heatmap visualizes for each curated clinical characteristic (rows) the availability in each dataset (columns). Red indicates that the corresponding characteristic is available for at least one sample in the dataset."}
.esetsStats <- function(esets) {
    res <- lapply(varLabels(esets[[1]]), function(covar) unlist(sapply(esets, 
        function(X) sum(!is.na(X[[covar]]))>0)))
    names(res) <- varLabels(esets[[1]])    
    do.call(rbind, res)
}

df.r <- .esetsStats(esets)
M <- as.matrix(apply(df.r,c(1,2),ifelse,0,1))
colnames(M) <- gsub("_eset$", "", colnames(M))
# no need to show the sample ids
M <- M[-(1:2),]
heatmap(M[nrow(M):1,],scale="none",margins=c(8,10),Rowv=NA)
```

## Summarizing the List of ExpressionSets

This example provides a table summarizing the datasets being used, and
is useful when publishing analyses based on curatedCRCData.
First, define some useful functions for this purpose:

```{r esetToTableFuns}
source(system.file("extdata", "summarizeEsets.R", package = "curatedCRCData"))
```

```{r esettable, echo=FALSE}
summary.table <- t(sapply(esets, getEsetData))
rownames(summary.table) <- sub("_eset", "", rownames(summary.table))
```

Optionally write this table to file, for example ( replace `myfile <- tempfile()` with something like `myfile <- "nicetable.csv"` )

```{r writeesettable}
(myfile <- tempfile())
write.table(summary.table, file=myfile, row.names=FALSE, quote=TRUE, sep=",")
```

```{r xtable, echo=FALSE, results='asis'}
library(knitr)
kable(summary.table, caption="Datasets provided by curatedCRCData.")
```

## For non-R users

If you are not doing your analysis in R, and just want to get some
data you have identified from the curatedCRCData manual, here is
a simple way to do it. For one dataset:

```{r simplygetdata, eval=FALSE}
library(curatedCRCData)
library(Biobase)
TCGA.COAD_eset <- curatedCRCData("TCGA.COAD_eset")
write.csv(exprs(TCGA.COAD_eset), file="TCGA.COAD_eset_exprs.csv")
write.csv(pData(TCGA.COAD_eset), file="TCGA.COAD_eset_clindata.csv")
```

Or for several datasets:

```{r simplyseveraldatasets, eval=FALSE}
data.to.fetch <- curatedCRCData(c("TCGA.COAD_eset", "GSE39582_eset"))
for (onedata in names(data.to.fetch)){
    write.csv(exprs(data.to.fetch[[onedata]]),
              file=paste(onedata, "_exprs.csv", sep=""))
    write.csv(pData(data.to.fetch[[onedata]]),
              file=paste(onedata, "_clindata.csv", sep=""))
}
```

## Session Info
```{r sessioninfo, echo=FALSE}
sessionInfo()
```